---
title: "Reproducible Research Project 1"
date: '2017-06-18'
output: html_document
---

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data
# Download file and unzip
```{r}
setwd("/Users/puru/Desktop/Reproducible Research/Week2/Course Project 1")
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if(!file.exists('data')) dir.create('data')
download.file(fileUrl, destfile = './data/activity.zip',method="curl")
unzip('./data/activity.zip', exdir = './data')
```

# Read csv file into data.frame
```{r}
dt <- data.table::fread(input = "data/activity.csv")
```

# Convert date factor to date
```{r}
tdt <- transform(dt, date=as.Date(date, format="%Y-%m-%d"))
```

# Remove the NA values and store in a separate structure
```{r}
dt1 <- dt[complete.cases(dt), ]
```

## What is mean total number of steps taken per day?

# Calculate the total number of steps taken per day
```{r}
steps_per_day <- aggregate(steps ~ date, dt1, sum)
```

# Create a histogram of no of steps per day
```{r}
hist(steps_per_day$steps, main = "Histogram of total number of steps per day", xlab = "Steps per day")
```

# Calculate the mean and median of the total number of steps taken per day
```{r}
round(mean(steps_per_day$steps))
median(steps_per_day$steps)
```

## What is the average daily activity pattern?
# Time series plot of the average number of steps taken
```{r}
avg_steps_per_interval <- aggregate(steps ~ interval, dt1, mean)
avg_steps_per_day <- aggregate(steps ~ date, dt1, mean)
plot(avg_steps_per_interval$interval, avg_steps_per_interval$steps, type='l', col=1, main="Average number of steps by Interval", xlab="Time Intervals", ylab="Average number of steps")
```

# The 5-minute interval that, on average, contains the maximum number of steps
# Identify the interval index which has the highest average steps
```{r}
interval_idx <- which.max(avg_steps_per_interval$steps)
```

# Identify the specific interval and the average steps for that interval
```{r}
print (paste("The interval with the highest avg steps is ", avg_steps_per_interval[interval_idx, ]$interval, " and the no of steps for that interval is ", round(avg_steps_per_interval[interval_idx, ]$steps, digits = 1)))
```

## Imputing missing values
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s)
```{r}
missing_value_act <- dt[!complete.cases(dt), ]
nrow(missing_value_act)
```

# Aggregate the steps per day with the imputed values
```{r}
steps_per_day_impute <- aggregate(steps ~ date, dt, sum)
```

# Draw a histogram of the imputed values
```{r}
hist(steps_per_day_impute$steps, main = "Histogram of total number of steps per day (IMPUTED)", xlab = "Steps per day")
```

# Compute the mean and median of the imputed value
# Calculate the mean and median of the total number of steps taken per day
```{r}
round(mean(steps_per_day_impute$steps))
median(steps_per_day_impute$steps)
```

## Are there differences in activity patterns between weekdays and weekends?
# Create a function to determine if the date is a weekday
```{r}
week_day <- function(date_val) {
    wd <- weekdays(as.Date(date_val, '%Y-%m-%d'))
    if  (!(wd == 'Saturday' || wd == 'Sunday')) {
        x <- 'Weekday'
    } else {
        x <- 'Weekend'
    }
    x
}
```

# Apply the week_day function and add a new column to dt dataset
```{r}
dt$day_type <- as.factor(sapply(dt$date, week_day))
```

# load the ggplot library
```{r}
library(ggplot2)
```

# Create the aggregated data frame by intervals and day_type
```{r}
steps_per_day_impute <- aggregate(steps ~ interval+day_type, dt, mean)
```

# Make a panel plot containing a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
plot <- ggplot(steps_per_day_impute, aes(interval, steps)) + geom_line(stat = "identity", aes(colour = day_type)) + theme_gray() +
facet_grid(day_type ~ ., scales="fixed", space="fixed") + labs(x="Interval", y=expression("No of Steps")) + ggtitle("No of steps Per Interval by day type")
print(plot)
```



